A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-03-27 analysis/paper/paper.Rmd

In this study, we performed a comprehensive evaluation for `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`, and `r ref("sfig", "extended_evaluation_overview")`). The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface, and the date of publication  (`r ref("table", "implementations")`). Only methods published before june 2017 are included in the current version of the evaluation. The evaluation comprised three core aspects: (i) source-code and literature-based characterisation of TI methods, (ii) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, and (iii) a quality control of the provided software and documentation.

In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. We then defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`). The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets came from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("sfig", "dataset_characterisation")` and `r ref("stable", "datasets")`). To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single-cell profiling experiment by matching the distributions of the synthetic data with real reference datasets (`r ref("sfig", "synthetic_data_generation")`).

```{r}
add_sfig(
  figure_file('dataset_characterisation.svg', '2-dataset_characterisation'),
  "dataset_characterisation",  
  "Characteristics of the real datasets used for the evaluation. Top: distributions of number of cells and genes (after filtering) and the date these datasets were published. Bottom: Diversity of technologies which were used to generate the dataset, the organism, whether we extracted a silver or a gold standard from the data, the type of dynamic process and the trajectory type present in the data."
)
```

```{r}
add_sfig(
  figure_file('synthetic_data_generation.svg', "2-dataset_characterisation/3-synthetic"),
  "synthetic_data_generation",
  "Workflow to generate the synthetic data. Starting from a state network, we extracted a regulatory network using real networks which is expected to induce such a set of state transitions when simulated. We simulated this network using a detailed model of gene regulation which includes the thermodynamics of transcription factor binding. Finally, we simulated the single-cell RNA-seq (scRNA-seq) experiment itself, by matching the distribution of expression and dropouts to a real reference dataset."
)
```

Finally, using both the benchmark results and quality control, we produced a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.

@@

We further investigated the robustness of our evaluation across datasets and dataset sources. Method performance was moderately to highly correlated (0.5-0.9) between real and synthetic datasets (`r ref("fig", "results_overview", "b")` and `r ref("sfig", "real_synthetic_comparison")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard. Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with most methods significantly outperforming the next method in the ranking `r ref("sfig", "performance_dataset_variability_combined")`. To further illustrate this, if we looked at the scores a random 10% subset of all datasets, the overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets. On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best. For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top`datasets on which it would be the top ranking method. Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large set of datasets, such as those included in our evaluation, is necessary to robustly compare the overall performance of methods.
```{r}
add_sfig(
  figure_file(
    "real_synthetic_comparison.svg",
    "11-evaluation_robustness"
  ), 
  "real_synthetic_comparison", 
  "Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
)
```

```{r}
add_sfig(
  figure_file(
    "performance_dataset_variability_combined.svg",
    "11-evaluation_robustness"
  ), 
  "performance_dataset_variability_combined", 
  "Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
)
```

```{r}
add_sfig(
  figure_file('metrics_comparison.svg', "5-optimise_parameters/7-metric_comparison"), 
  "metric_comparison", 
  "Comparison of method performance between different metrics across trajectory types. For each row, a different pairwise comparison is given between two metrics."
)
```

### Trajectory types and topology complexity

During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error was produced, a zero score was returned for that execution. If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”. Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g. Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions. Methods that do not scale well with respect to the number of cells or genes will exceed the time or memory limits for the largest datasets (reCAT, SCOUP, StemID). For a few methods, the time or memory limits were exceeded too often, making the benchmarking results uncomparable to those of other methods (SCIMITAR, Ouija, Pseudogp).

### Trajectory types and topology complexity

In most cases, the methods which were specifically designed to handle a particular trajectory type, also performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`). These methods had typically better edge flip scores - as can be expected - and RF MSE scores, compared to methods not able to handle the particular trajectory type. However, the correlation score typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories.
To further investigate the effect of trajectory complexity on performance, we divided them in two groups: linear methods (restricted to linear and cyclic topologies) and non-linear methods. While there were no significant differences in performance on linear datasets, non-linear methods had significantly higher edge flip scores but lower correlation scores on datasets containing more complex trajectory types  (`r ref("sfig", "linear_split_comparison")`). Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms. A combination of the ordering methods from the top linear methods, combined with the topology inference from top non-linear methods, could therefore be a possibility for future research.

```{r}
add_sfig(
  figure_file('trajtype_handle_comparison.svg', "8-compare_topology"), 
  "trajtype_handle_comparison", 
  "Comparison between the performance of methods able to handle a trajectory type with those who don't. Shown is the performance on those datasets containing a particular trajectory type (top), according to several metrics (right)."
)
```

```{r}
add_sfig(
  figure_file('linear_split_comparison.svg', "8-compare_topology"), 
  "linear_split_comparison", 
  "Comparison of the performance between linear and non-linear methods. Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods. Their performance is shown on linear and non-linear datasets according to several performance metrics (right). NS: Not significant, \\*: p-value < 0.05, \\*\\*: p-value < 0.001, \\*\\*\\*: p-value < 0.0001"
)
```

Despite their similar overall performance, the topologies predicted by the top methods differed considerably. Trajectories detected using the default parameters of slingshot and cellTree tended to be simpler, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`). Monocle DDRTree, for example frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`). Trajectories generated by Monocle DDRTree (at default parameters) tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets. Indeed, when we assessed how often a method can infer the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies. Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases (`r ref("sfig", "topology_sensitivity")`). Inferring the correct topology without parameter tuning, is therefore still an open challenge. Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the end user, to either optimise the parameters or to choose the method with which the output best fits the user's expert knowledge.

@@

----

```{r}
add_sfig(
  figure_file('topology_sensitivity.svg', "8-compare_topology"), 
  "topology_sensitivity", 
  "Sensitivity of each TI method in predicting the correct trajectory topology. Shown are the % of datasets where a method is able to predict the exact correct trajectory topology, depending on the trajectory type present in the data. When a method is able to predict a particular trajectory type, the box is surrounded by a black border."
)
```

### Effect of prior information

In the current version of the evaluation, we only provided prior information when a method required it. We did not observe a major difference in method performance between methods which did and did not receive prior information (`r ref("sfig", "required_priors_performance_comparison")`). Rather, methods which received prior information were on average positioned in the middle of the ranking.  Furthermore, we could not find any dataset where methods which received prior information performed significantly better than other methods.

```{r}
add_sfig(
  figure_file('required_priors_performance_comparison.svg', "10-prior_information"), 
  "required_priors_performance_comparison", 
  "Comparison of method performance depending on whether the method required a certain prior or not. The top performing method which requires a particular type of prior information is labelled."
)
```

### Algorithm components

We assessed whether the components of an algorithm could  be predictive of a method’s performance, using both random forest classification and statistical testing. Methods which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance (`r ref("sfig", "method_components_scores")` and `r ref("stable", "component_importance")`). On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.

```{r}
add_sfig(
  figure_file("method_components_scores.svg","6-developer_guidelines"), 
  "method_components_scores", 
  "Score of methods containing particular algorithmic components. Shown are the algorithmic components present in four or more methods, and whether the performance of methods containing a certain component was significantly higher than all the other methods. NS: Not significant. *: corrected p-value < 0.05 "
)
```

@@

----

```{r}
add_sfig(
  figure_file("qc_over_time.svg","4-method_characterisation"), 
  "qc_over_time", 
  "Quality control scores and number of citations over time."
)
```
We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")` and `r ref("sfig", "qc_performance_comparison")`). We could also not find any quality aspect which was significantly associated with method performance.

```{r}
add_sfig(
  figure_file(
    "qc_performance_comparison.svg",
    "5-optimise_parameters/6-qc_performance_comparison"
  ), 
  "qc_performance_comparison", 
  "Comparison between method quality control scores and overall method performance."
)
```

@@

# Methods

@@

```{r}
add_sfig(
  figure_file("implementation_components.svg","4-method_characterisation"), 
  "implementation_components", 
  "Common algorithmic components shared by different TI methods. Most components can be categorised into 4 categories: clustering (pink), dimensionality reduction (orange), generative models (green) and graph building (blue). a) Wordcloud of the common components. b) Number of times a component was shared between methods. c) Number of methods containing a particular category."
)
```

@@

## Real datasets
We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")` and `r ref("sfig", "dataset_characterisation")`). The scripts to download and process these datasets will be made available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalised and filtered using a common pipeline, discussed later. We determined a reference standard for every dataset using labelling provided by the author's, and classified the standards into gold and silver based on whether this labelling was determined by the expert using the expression (silver standard) or using other external information (such as FACS or the origin of the sample, gold standard) (`r ref("stable", "datasets")`).

@@

Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment (`r ref("sfig", "synthetic_data_generation")`). At every step, we took great care to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable. For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.

@@

To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters (`r ref("sfig", "modulenets")`). Two module networks generated linear trajectories (linear and linear long), two generated simple forks (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple undirected graphs (bifurcating loop and bifurcating convergence).

@@

```{r}
add_sfig(
  figure_file("modulenets.svg", experiment_id = "2-dataset_characterisation/3-synthetic"),
  "modulenets", 
  "Module networks, milestone networks and the module dynamics for each of the different types of synthetic data", 
  15, 
  5
)
```

@@

```{r}
add_sfig(
  figure_file("score_aspects_v1.svg", experiment_id="3-score_rules/5-score_aspects"), 
  "score_aspects", 
  "Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.",
  15, 
  8
)
```

@@

### Method execution
Each execution of a method on a dataset was performed in a separate task as part of a gridengine job. Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 @ 2.40GHz, and was provided a maximum of 32GB in memory and 6 hours of wall time. One R session was started for each task, with the environment variable `R_MAX_NUM_DLLS` set to 500. The `base::set.seed` was overridden in order to prevent stochastic TI methods from pretending to be deterministic and/or robust. Timings of methods were measured for different steps along the executions, including preprocessing, postprocessing, each of the different metrics, and the method itself.

@@

## Trajectory types
We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria. These trajectory types start from the most general trajectory type, a disconnected directed graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical (`r ref("sfig", "trajectory_type_tree", "a")`). For every directed trajectory type, a corresponding undirected trajectory type can also be defined (`r ref("sfig", "trajectory_type_tree", "b")`). In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with the only exception being DPT (limited to bifurcating trajectories).

```{r}
add_sfig(
  figure_file("trajectory_type_trees.svg","2-dataset_characterisation/2-trajectory_types"), 
  "trajectory_type_tree", 
  "Different trajectory types can be classified in a directed acyclic graph structure, in which certain types are a generalisation of other trajectory types. a) The graph structure for undirected and directed trajectory types. Shown are the topological properties which change when going from a particular trajectory type to a less complex type. b) Combined structure for directed and undirected trajectory types.", 
  15, 
  10
)
```

# Supplementary Material
```{r}
add_sfig(
  figure_file("methods_timeline.svg", "4-method_characterisation"),
  "methods_over_time", 
  "New TI methods over time. a) Number of methods published or in preprint. b) Number of methods fixing the trajectory topology, either by design (fixed) or through user parameters (parameter). c) Number of methods which can handle a particular type of trajectory."
)
```

```{r, results='asis' }
add_stable(
  "implementations",
  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
  paste0("Overview of the trajectory inference methods included in this study, and several characteristics thereof. This table will be continuously updated ", url("https://goo.gl/FnsTxG", "online"), ".")
)
```

```{r "supplementary", echo=FALSE, results='asis'}
source("analysis/paper/sfigs_stables.R")
```

# Colophon

This report was generated on `r Sys.time()` using `r devtools:::platform_info()$version` and the following packages:
```{r colophon, cache = FALSE}
devtools::session_info()$packages %>% knitr::kable()
```

>>>>

In this study, we performed a comprehensive evaluation of  `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`). The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface (`r ref("table", "implementations")`). The evaluation comprised three core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, (ii) a quality control of the provided software and documentation, and (iii) combining both evaluations to create a set of practical guidelines for users.

@@

We further investigated the robustness of our evaluation across datasets and dataset sources. Method performance was moderately to highly correlated (0.5-0.9 depending on the trajectory type) between real and synthetic datasets (`r ref("sfig", "robustness_comparison", "a")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard. Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with a majority of the methods significantly outperforming the next method in the ranking (`r ref("sfig", "robustness_comparison", "b")`). To further illustrate this, we looked at the scores in a randomly selected 10% subset of all datasets. The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets. On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best. For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method. Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large set of datasets, such as those included in our evaluation, is necessary to robustly compare the overall performance of methods.

## Relationships between method characteristics to performance

<!-- Inverse relation between edge flip and correlation -->
In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`). These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type. However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories. To further investigate the effect of trajectory complexity on performance, we divided the methods in two groups: non-branching methods (restricted to linear and cyclic topologies) and branching methods. While there were no significant differences in performance on linear datasets, branching methods produced significantly a better topology, but a less accurate cellular ordering on datasets containing at least one branching point (`r ref("sfig", "linear_split_comparison")`). Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms. A combination of the ordering methods from the top linear methods, combined with the topology inference from top non-linear methods, could therefore be a possibility for future research.

<!-- Topological complexity versus performance -->
Despite their similar overall performance, the topologies predicted by the top methods differed considerably. Trajectories detected by slingshot and cellTree tended to contain less branches, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`). Monocle DDRTree for example, frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`). Trajectories generated by Monocle DDRTree tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets. Indeed, when we assessed how often a method inferred the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies. Nonetheless, the overall accuracy of topology prediction was very low, with slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases. Inferring the correct topology without parameter tuning is therefore still an open challenge. Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the user, to either optimise the parameters or to choose the method whose output best fits the user's expert knowledge.

@@

<!-- Additional factors which could influence performance -->
Apart from the trajectory types, we also investigated several other factors which could influence the performance of a method. Although we provided prior information to those methods which require it, we did not observe any major difference in method performance between these methods and methods that did not receive any prior information (`r ref("fig", "results_overview", "b")`). When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as slingshot and SCORPIUS), k-means (which include slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance ( `r ref("stable", "component_importance")`). On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.

@@

We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")`). We could also not find any quality aspect which was significantly associated with method performance.

@@

<!-- The context -->
Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data. Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017]. It is therefore critical that these methods, now reaching `r sum(implementations$is_ti, na.rm=TRUE)`, are evaluated to guide users in their choice. In this study, we presented a large scale evaluation of the performance and code of  `r sum(methods$evaluated)` TI methods. When comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art. We believe that the benchmarking presented in this study will pave the way to the next series improvements in the field of trajectory inference.

# Online Methods

@@

@@

#### cellTree

## Real datasets
We gathered a list of real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`). The scripts to download and process these datasets will be made available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalised and filtered using a common pipeline, discussed later. We determined a reference standard for every dataset using labelling provided by the author's, and classified the standards into gold and silver based on whether this labelling was determined by the expert using the expression (silver standard) or using other external information (such as FACS or the origin of the sample, gold standard) (`r ref("stable", "datasets")`).

@@

Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment. At every step, we took great care to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable. For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.

@@

To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters. Two module networks generated linear trajectories (linear and linear long), two generated simple forks (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple undirected graphs (bifurcating loop and bifurcating convergence).

@@

@@

```{r}
add_sfig(
  figure_file("score_aspects_v2.svg", experiment_id="3-score_rules/5-score_aspects"), 
  "score_aspects", 
  "Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.",
  15, 
  8
)
```

@@

### Method execution
Each execution of a method on a dataset was performed in a separate task as part of a gridengine job. Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 @ 2.40GHz, and one R session was started for each task. During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error was produced, a zero score was returned for that execution. If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”. The environment variable `R_MAX_NUM_DLLS` set to 500. The `base::set.seed` was overridden in order to prevent stochastic TI methods from pretending to be deterministic and/or robust. Timings of methods were measured for different steps along the executions, including preprocessing, postprocessing, each of the different metrics, and the method itself.

@@

## Trajectory types
We classified all possible trajectory topologies into distinct trajectory types, based on topological criteria. These trajectory types start from the most general trajectory type, a disconnected directed graph, and move down (within a directed acyclic graph structure), progressively becoming more simple until the two basic types: linear and cyclical. A disconnected directed graph is a graph in which only one edge can exist between two nodes. A directed graph is a disconnected graph in which all nodes are (indirectly) connected. A directed acyclic graph is a graph containing no cycles. A rooted tree is a directed acyclic graph containing no convergences (no nodes with in-degree higher than 1). A rooted binary tree is a rooted tree with every node having maximally two outgoing edges. A convergence is a directed acyclic graph in which only one node has a degree larger than one and this same node has an indegree of one. A multifurcation is a rooted tree in which only one node has a degree larger than one. A bifurcation is a multifurcation in which only one node has a degree equal to 3. A directed linear graph is a graph in which no node has a degree larger than 3. Finally, a directed cycle is a directed graph in which no node has a degree larger than 3 but contains one cycle. For every directed trajectory type, a corresponding undirected trajectory type can also be defined, with directed acyclic graphs and directed graphs merging to undirected graph, and convergence and bifurcation merging to a simple fork. In most cases, a method which was able to detect a complex trajectory type, was also able to detect less complex trajectory types, with the only exception being DPT (limited to bifurcating trajectories).

# Supplementary Material

```{r}
add_sfig(
  figure_file(
    "robustness.svg",
    "11-evaluation_robustness"
  ), 
  "robustness_comparison", 
  "**Comparative analysis between various facets of the results.** a) Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation. b) Violin plots of the overall performance. For each method, the arch indicates next method in the ranking over which one has significant improvement in performance over the other (top; pairwise wilcoxon test corrected for multiple testing; p-value < 0.05)."
)
```

```{r, results='asis' }
add_stable(
  "implementations",
  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
  "Overview of the trajectory inference methods included in this study, and several characteristics thereof."
)
```

1) method performance highly depends on trajectory type
2) methods which are specific to a certain trajectory type generally perform better on datasets containing these trajectory types
3) the complexity of the topologies returned by the different methods is highly variable 
4) prior information doesn't increase performance in our settings
